---
title: "R Notebook"
output: html_notebook
---


```{r}
# load libaries
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(DBI)
library(RSQLite)
```

### Data analysis & exploration
```{r}

# load data tables
bed_fact_tab <- read.csv("bed_fact.csv")
bed_type_tab <- read.csv("bed_type.csv")
business_tab <- read.csv("business.csv")

# data summary
head(bed_fact_tab)
summary(bed_fact_tab)

head(bed_type_tab)
summary(bed_type_tab)

head(business_tab)
summary(business_tab)

# check outliers outliers 

boxplot(bed_fact_tab$license_beds, main="License Beds")
boxplot(bed_fact_tab$census_beds, main="Census Beds")
boxplot(bed_fact_tab$staffed_beds, main="Staffed Beds")

# remove outliers using interquatile range 

remove_outliers <- function(x) {
  Q1 <- quantile(x, 0.25)
  Q3 <- quantile(x, 0.75)
  IQR <- Q3 - Q1
  lower_bound <- Q1 - 1.5 * IQR
  upper_bound <- Q3 + 1.5 * IQR
  return(ifelse(x >= lower_bound & x <= upper_bound, x, NA))
}
# Remove outliers from license_beds, census_beds, and staffed_beds columns
bed_fact_tab_cleaned <- bed_fact_tab %>%
  mutate(
    license_beds = remove_outliers(license_beds),
    census_beds = remove_outliers(census_beds),
    staffed_beds = remove_outliers(staffed_beds)
  )

# Remove rows with NA values
bed_fact_tab_cleaned <- bed_fact_tab_cleaned %>%
  drop_na()

# Check the shape of the original and cleaned dataframe
cat("Shape of original dataframe (bed_fact_df):\n")
cat(dim(bed_fact_tab), "\n")
cat("Shape of cleaned dataframe (bed_fact_df_cleaned):\n")
cat(dim(bed_fact_tab_cleaned), "\n")

# Visualize boxplots after removing outliers
boxplot(bed_fact_tab_cleaned$license_beds, main="License Beds (No Outliers)")
boxplot(bed_fact_tab_cleaned$census_beds, main="Census Beds (No Outliers)")
boxplot(bed_fact_tab_cleaned$staffed_beds, main="Staffed Beds (No Outliers)")

```

#### 1. Identify the dimensions from each dimension table

For the bed_type table:  

- Fact: None     
- Dimensions: bed_id, bed_code, and bed_desc   

For the business_csv table:

- Dimensions: ims_org_id, business_name, and bed_cluster_id.  

#### 2. Identify the Facts variables from the single Fact Table 

For bed_fact table: 

- Facts: license_beds, census_beds and staffed_beds.    

### 3a. Analysis for Leadership
```{r}

# Create SQLite connection and register data frames
con <- dbConnect(RSQLite::SQLite(), dbname = ":memory:")
dbWriteTable(con, "bed_fact_tab", bed_fact_tab)
dbWriteTable(con, "bed_type_tab", bed_type_tab)
dbWriteTable(con, "business_tab", business_tab)
dbWriteTable(con, "bed_fact_tab_cleaned", bed_fact_tab_cleaned)
dbWriteTable(con, "icu_sicu_beds", icu_sicu_beds)



# Identify which hospitals have an ICU bed  or a SICU bed.

query <- "
  SELECT bed_fact_tab_cleaned.*, 
    bed_type_tab.bed_code,
    bed_type_tab.bed_desc
  FROM bed_fact_tab_cleaned
  LEFT JOIN bed_type_tab ON bed_fact_tab_cleaned.bed_id = bed_type_tab.bed_id
"
icu_sicu_beds <- dbGetQuery(con, query)

icu_sicu_beds <- icu_sicu_beds[icu_sicu_beds$bed_id %in% c(4, 15), ]
icu_sicu_beds


# Perform left join between business_tab and icu_sicu_beds
query <- "
  SELECT business_tab.*,
    icu_sicu_beds.bed_id,
    license_beds,
    census_beds,
    staffed_beds,
    bed_code,
    bed_desc
  FROM business_tab 
  LEFT JOIN icu_sicu_beds ON business_tab.ims_org_id = icu_sicu_beds.ims_org_id
"
summary_data_icu_sicu_beds <- dbGetQuery(con, query)

# Remove duplicates 
summary_data_icu_sicu_beds <- summary_data_icu_sicu_beds[!duplicated(summary_data_icu_sicu_beds$business_name), ]

dbWriteTable(con, "summary_data_icu_sicu_beds", summary_data_icu_sicu_beds)

# 1.License beds:
query <- "
SELECT business_name, ttl_license_beds
FROM summary_data_icu_sicu_beds
ORDER BY ttl_license_beds DESC
LIMIT 10;
"
license_beds_hospitals <- dbGetQuery(con, query)
license_beds_hospitals

# 2.Census beds:
query <- "
SELECT business_name, ttl_census_beds
FROM summary_data_icu_sicu_beds
ORDER BY ttl_census_beds DESC
LIMIT 10;
"
census_beds_hospitals <- dbGetQuery(con, query)
census_beds_hospitals

# 3.Staffed beds:

query <-"SELECT business_name, ttl_staffed_beds
FROM summary_data_icu_sicu_beds
ORDER BY ttl_staffed_beds DESC
LIMIT 10;
"
staffed_beds_hospitals <- dbGetQuery(con, query)
staffed_beds_hospitals

# Close the SQLite connection
dbDisconnect(con)
```

### 3b. Interpretation of Findings 

- Florida State Hospital has the highest number of licensed beds among all hospitals.  

- Patton State Hospital has the highest number of census beds and staffed beds.   

- Florida State Hospital, Jackson Memorial Hospital, and The Cleveland Clinic Foundation are among the top 10 
hospitals with the highest number of licensed beds, census beds, and staffed beds.  

### 4a. Drill down investigation 
```{r}
# Create SQLite connection and register data frames
con <- dbConnect(RSQLite::SQLite(), dbname = ":memory:")
dbWriteTable(con, "bed_fact_tab", bed_fact_tab)
dbWriteTable(con, "bed_type_tab", bed_type_tab)
dbWriteTable(con, "business_tab", business_tab)
dbWriteTable(con, "bed_fact_tab_cleaned", bed_fact_tab_cleaned)
dbWriteTable(con, "hospitals_with_both_icu_sicu", hospitals_with_both_icu_sicu)
dbWriteTable(con, "summary_data_both_icu_sicu", summary_data_both_icu_sicu)

# Identify hospitals with both ICU and SICU beds
query <- "
  SELECT ims_org_id
  FROM bed_fact_tab_cleaned
  WHERE bed_id IN (4, 15)
  GROUP BY ims_org_id
  HAVING COUNT(DISTINCT bed_id) = 2
"
hospitals_with_both_icu_sicu <- dbGetQuery(con, query)

# Perform left join between business_tab and hospitals_with_both_icu_sicu
query <- "
  SELECT business_tab.*,
    bed_fact_tab_cleaned.bed_id,
    license_beds,
    census_beds,
    staffed_beds
  FROM business_tab 
  LEFT JOIN bed_fact_tab_cleaned ON business_tab.ims_org_id = bed_fact_tab_cleaned.ims_org_id
  WHERE bed_fact_tab_cleaned.ims_org_id IN (SELECT ims_org_id FROM hospitals_with_both_icu_sicu)
"
summary_data_both_icu_sicu <- dbGetQuery(con, query)

# Remove duplicates 
summary_data_both_icu_sicu <- summary_data_both_icu_sicu[!duplicated(summary_data_both_icu_sicu$business_name), ]

# Summary report for License beds
query <- "
  SELECT business_name, SUM(license_beds) AS total_license_beds
  FROM summary_data_both_icu_sicu
  GROUP BY business_name
  ORDER BY total_license_beds DESC
  LIMIT 10
"
license_beds_both_icu_sicu <- dbGetQuery(con, query)
license_beds_both_icu_sicu
# Summary report for Census beds
query <- "
  SELECT business_name, SUM(census_beds) AS total_census_beds
  FROM summary_data_both_icu_sicu
  GROUP BY business_name
  ORDER BY total_census_beds DESC
  LIMIT 10
"
census_beds_both_icu_sicu <- dbGetQuery(con, query)
census_beds_both_icu_sicu
# Summary report for Staffed beds
query <- "
  SELECT business_name, SUM(staffed_beds) AS total_staffed_beds
  FROM summary_data_both_icu_sicu
  GROUP BY business_name
  ORDER BY total_staffed_beds DESC
  LIMIT 10
"
staffed_beds_both_icu_sicu <- dbGetQuery(con, query)
staffed_beds_both_icu_sicu

# Close the SQLite connection
dbDisconnect(con)

```

### 4b. Final recommendation

Based on the analysis conducted in steps 3a and 4a, **The Cleveland Clinic Foundation and Baptist Health Medical Center Little Rock** stand  as strong candidates for the pilot intervention program for the following reasons:


1. The Cleveland Clinic Foundation: as one of the top 10 hospitals with the highest number of licensed beds, census beds, and staffed beds, The Cleveland Clinic Foundation demonstrates a strong infrastructure and capability to handle a significant volume of patients. 

2. Baptist Health Medical Center Little Rock: Despite not being among the top 10 hospitals in the initial analysis, Baptist Health Medical Center Little Rock stands out in step 4a with the highest total number of licensed beds and staffed beds. This indicates a substantial capacity to accommodate patients and provide adequate staffing for their care needs. Additionally, its location in Little Rock may offer strategic advantages in terms of accessibility and community impact.

By selecting these two hospitals, the pilot intervention program can benefit from their combined strengths in infrastructure, capacity, and expertise, ensuring a comprehensive approach to healthcare delivery and maximizing the potential for success.

